---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Turning off scientific notation and calling a package for rounding inline text.
```{r,echo=TRUE}
options(scipen=999)
library(broman)
```

## Loading and preprocessing the data

1. Load the data (i.e. read.csv())
```{r,echo=TRUE}
setwd("~/Coursera/5 - Reproducible Research/RepData_PeerAssessment1")
filename <- unzip("activity.zip")
pers.act <- read.csv("activity.csv", header = TRUE, sep = ',')
```

2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r,echo=TRUE}
pers.act$date <- as.Date(pers.act$date)
head(pers.act)
str(pers.act)
```

## What is mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day
```{r,echo=TRUE}
steps.day <- aggregate(steps ~ date, data = pers.act, sum, na.rm = TRUE)
head(steps.day)
library(ggplot2)
ggplot(steps.day, aes(x=steps)) +
       geom_histogram(binwidth=1000) +
       labs(title = "Total Steps Taken Each Day", 
            x = "Steps per Day",
            y = "Frequency")
```

2. Calculate and report the mean and median total number of steps taken per day
```{r,echo=TRUE}
mean.steps <- mean(steps.day$steps)
mean.steps
median.steps <- median(steps.day$steps)
median.steps
```
The mean total number of steps taken per day is `r round(mean.steps, 2)`, whereas the median total number of steps taken per day is `r median.steps`. 

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r,echo=TRUE}
pers.int <- aggregate(steps ~ interval, data = pers.act, mean, na.rm = TRUE)
head(pers.int)
ggplot(pers.int, aes(x=interval, y=steps)) +
       geom_line(color="black") +
       labs(title = "Average Steps Taken per 5-minute Interval", 
            x = "5-minute Interval",
            y = "Average Steps Taken")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r,echo=TRUE}
max.int <- pers.int$interval[which.max(pers.int$steps)]
max.int
```

On average, the `r max.int`th interval contains the maximum number of steps.

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r,echo=TRUE}
miss.row <- sum(is.na(pers.act))
miss.row
```

There are `r miss.row` rows with NAs in the dataset.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

I will impute missing values for steps with the mean number of steps with the average number of steps for a given 5-minute interval.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r,echo=TRUE}
pers.act.imp <- pers.act
pers.act.imp.na <- is.na(pers.act.imp$steps)
imp.values <- tapply(pers.act.imp$steps, pers.act.imp$interval, mean, na.rm=TRUE, simplyfy=TRUE)
pers.act.imp$steps[pers.act.imp.na] <- 
    imp.values[as.character(pers.act.imp$interval[pers.act.imp.na])]
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r,echo=TRUE}
steps.day.imp <- aggregate(steps ~ date, data = pers.act.imp, sum, na.rm=TRUE)
head(steps.day.imp)
ggplot(steps.day.imp, aes(x=steps)) +
       geom_histogram(binwidth=1000) +
       labs(title = "Total Steps Taken Each Day with Missing Values Imputed", 
            x = "Steps per Day",
            y = "Frequency")
mean.steps.imp <- mean(steps.day.imp$steps)
mean.steps.imp
median.steps.imp <- median(steps.day.imp$steps)
median.steps.imp
```

Imputing missing values with the average number of steps for a given 5-minute interval changes the median, such that the mean and median are now equal to the same value, `r round(mean.steps.imp, 2)`.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r,echo=TRUE}
pers.act.imp$dow <- weekdays(pers.act.imp$date)
pers.act.imp$dow <- ifelse(pers.act.imp$dow == "Saturday" | 
                           pers.act.imp$dow == "Sunday", "weekend", "weekday")
head(pers.act.imp)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r,echo=TRUE}
pers.int.week <- aggregate(steps ~ interval + dow, data = pers.act.imp, mean, na.rm = TRUE)
head(pers.int.week)
ggplot(pers.int.week, aes(x=interval, y=steps, color = dow)) +
       geom_line() +
       facet_wrap(~dow, ncol = 1, nrow=2) +
       labs(title = "Average Steps Taken per Interval on Weekdays vs. Weekends", 
            x = "5-minute Interval",
            y = "Average Steps Taken",
            color = "Legend")
```